\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-03-10 k-planar.tex
first draft of proof in appendix complete

\begin{lem}
Say $G_0$ is a 2-connected plane multigraph. Let $T$ be a spanning tree of $G_0$ rooted at some vertex $r$ on the boundary of the outerface of $G_0$. Let $k\in\{1,2,3\}$ and let $C = [A_1,\dots,A_k]$ be a cycle in $G_0$ with $r$ in the exterior of $C$, where $A_1, \dots, A_k$ are bipods in $G_0$. Let $G$ be the plane graph consisting of all the edges and vertices of $G_0$ in $C$ and the interior of $C$. Then $G$ has a partition $\PP$ into tripods with respect to $T$ such that:
\begin{compactitem}
\item $P_1,\dots,P_k\in\PP$,
\item the quotient $G / \PP$ is planar and has a tree-decomposition $\mathcal{T}$ of width at most 3 such that:
\begin{compactitem}
	\item some bag of $\mathcal{T}$ contains $P_1,\dots,P_k$, and 
	\item for every face $F$ of $G$ there is a bag $B$ of $\mathcal{T}$ such that $V(F)\subseteq \bigcup_{X\in B} X$.
\end{compactitem}
\end{compactitem}
\end{lem}

\begin{proof}
	\note{DW}{Can we always assume $k=3$ by allowing $A_i=\emptyset$? That would clean it up.}

@@

Consider $j\in\{1,\dots,m\}$. Let $G_j$ be the subgraph of $G$ consisting of all the edges and vertices of $G$ in $Z_j$ and the interior of $Z_j$. The outerface of $G_j$ is $Z_j$. Observe that for some $i\in\{1,\dots,k\}$, we have
$Z_j=[R_{i,j}^+,R_{i+1,j}^-,S_{i,j}]$, where 
$R_{i,j}^+$ is some subpath of $R_i^+$,
$R_{i+1,j}^-$ is some subpath of $R_{i+1}^-$, and
$S_{i,j}$ is some subpath of $S_i$.
Since $R_i^+$, $R_{i+1}^-$ and $S_i$ are bipods, 
$R_{i,j}^+$, $R_{i+1,j}^-$ and $S_{i,j}$ are bipods. 
By construction, $v_{i-1}$ is not in $G_j$. 
Thus $V(G_j)\setminus V(Z_j) \subsetneq V(G)\setminus V(C)$, and we may apply induction to $G_j$.  Thus $G_j$ has a partition $\PP_j$ into tripods with respect to $T$ such that:
\begin{compactitem}
	\item $R_{i,j}^+, R_{i+1,j}^-, S_{i,j}\in\PP_j$,
	\item the quotient $G_j / \PP_j$ is planar and has a tree-decomposition $\mathcal{T}_j$ of width at most 3 such that: 
	\begin{compactitem}
		\item some bag $B_j$ of $\mathcal{T}_j$ contains $R_{i,j}^+, R_{i+1,j}^-, S_{i,j}$, and 
		\item for every face $F$ of $G_j$ there is a bag $B$ of $\mathcal{T}_j$ such that $V(F)\subseteq \bigcup_{X\in B} X$.
	\end{compactitem}
\end{compactitem}
\end{lem}

@@

We now construct the desired partition $\PP$ of $G$. 
Initialise $\PP:= \{P_1,\ldots,P_k\}$. Then add each non-empty $Q_i'$ to $\PP$. 
Now for each non-degenerate $F_i$, each path in  $\PP_i$ is either an \emph{external} path (that is, fully contained in $F_i$) or is an \emph{internal path} with none of its vertices in $F_i$. 
Add all the internal paths of  $\PP_i$ to $\PP$. 
By construction, $\PP$ partitions $V(G)$ into vertical paths in $T$ and $\PP$ contains $P_1,\ldots,P_k$.

Let $H:=G/\PP$. Next we exhibit the desired tree-decomposition $(B_x:x\in V(J))$ of $H$. 
Let $J$ be the tree obtained from the disjoint union of $J_i$, taken over the $i\in\{1,2,3\}$ such that $F_i$ is non-degenerate, by adding one new node $u$ adjacent to each $u_i$.
(Recall that $u_i$ is the node of $J_i$ for which the bag $B^i_{u_i}$ contains the vertices of $H_i$ corresponding to the paths that form $F_i$.)\ 
Let the bag $B_u$ contain all the vertices of $H$ corresponding to $P_1,\ldots,P_k,Q'_1,Q'_2,Q'_3$. 
For each non-degenerate $F_i$, and for each node $x\in V(J_i)$, initialise $B_x:= B^i_x$. 
Recall that vertices of $H_i$ correspond to contracted paths in  $\PP_i$.
Each internal path in $\PP_i$ also lies  in $\PP$.
Each external path $P$ in $\PP_i$ is a subpath of $P_j$ for some $j\in\{1,\dots,k \}$ or is one of the paths among $Q'_1, Q'_2, Q'_3$. 
For each such path $P$, for every $x\in V(J)$, in bag $B_x$, 
replace each instance of the vertex of $H_i$ corresponding to $P$ by the vertex of $H$ corresponding to the path among $P_1,\ldots, P_k, Q'_1,\ldots,Q'_3$ 
that contains $P$.
This completes the description of $(B_x : x\in V(J))$. 
By construction, $|B_x|\leq 9$ for every $x\in V(J)$.

First we show that for each vertex $a$ in $H$, the set $X:=\{x\in V(J) : a\in B_x\}$ forms a subtree of $J$.
If $a$ corresponds to a path distinct from $P_1,\ldots,P_k,Q'_1,Q'_2,Q'_3$ then $X$ is fully contained in $J_i$ for some $i\in\{1,2,3\}$.
Thus, by induction $X$ is non-empty and connected in $J_i$, so it is in $J$.
If $a$ corresponds to $P$ which is one of the paths among $P_1,\ldots,P_k,Q'_1,Q'_2,Q'_3$ then 
$u\in X$ and whenever $X$ contains a vertex of $J_i$ it is because some external path of $\PP_i$ was replaced by $P$.
In particular, we would have $u_i \in X$ in that case. 
Again by induction each $X\cap J_i$ is connected and since $uu_i\in E(T)$, we conclude that $X$ induces a (connected) subtree of $J$.

Finally we show that, for every edge $ab$ of $H$, there is a bag $B_x$ that contains $a$ and $b$. 
If $a$ and $b$ are both obtained by contracting any of $P_1,\ldots,P_k,Q'_1,Q'_2,Q'_3$, 
then $a$ and $b$ both appear in $B_u$.  
If $a$ and $b$ are both in $H_i$ for some $i\in\{1,2,3\}$, 
then some bag $B^i_x$ contains both $a$ and $b$. 
Finally, when $a$ is obtained by contracting a path $P_a$ in $G_i-V(F_i)$ and $b$ is obtained by contracting a path $P_b$ not in $G_i$, 
then  the cycle $F_i$ separates $P_a$ from $P_b$ so the edge $ab$ is not present in $H$. 
This concludes the proof that $(B_x:x\in V(J))$ is the desired tree-decomposition  of $H$.

We construct $\PP$ as before. Initialise $\PP:=\{P_1,\ldots,P_k,Y\}$. Then, for $i\in \{1,2,3\}$, each tripod in  $\PP_i$ is either fully contained in $F_i$ or it is \emph{internal} with none of its vertices in $F_i$. Add all these internal tripods in  $\PP_i$ to $\PP$. By construction, $\PP$ partitions $V(G)$ into tripods. The graph $H:= G/ \PP$ is planar since $G$ is planar and each tripod in $\PP$ induces a connected subgraph of $G$.

Next we produce the tree-decomposition $(B_x:x\in V(J))$ of $H$ that satisfies the requirements of the lemma.  Let $J$ be the tree obtained from the disjoint union of $J_1$, $J_2$ and $J_3$ by adding one new node $u$ adjacent to $u_1$, $u_2$ and $u_3$. Let $B_u$ be the set of at most four vertices of $H$ corresponding to $Y,P_1,\dots,P_k$. For $i\in\{1,2,3\}$ and for each node $x\in V(J_i)$, initialise $B_x:= B^i_x$.

As in  the proof of \cref{NearTriang},  the resulting structure,  $(B_x:x\in V(J))$, is not yet a tree-decomposition of $H$ since some bags may  contain vertices of $H_i$ that are not necessarily vertices of $H$. Note that unlike in  \cref{NearTriang} this does not only include elements of $\PP_i$  that are contained in $F$. 
In particular, $I_i$ is also not an element of $\PP$ and thus does not correspond to a vertex of $H$.   
We remedy this  as follows. For $x\in V(J)$, in bag $B_x$, replace each instance of the vertex of $H_i$ corresponding to $I_i$ by the vertex of $H$ corresponding to $Y$. 
Similarly, by construction, $R_i^+$ is a subgraph of $P_{\alpha_i}$ for some $\alpha_i\in \{1,\dots,k\}$. 
For $x\in V(J)$, in bag $B_x$, replace each instance of the vertex of $H_i$ corresponding to $R_i^+$ 
by the vertex of $H$ corresponding to $P_{\alpha_i}$. 
Finally, $R_{i+1}^-$ is a subgraph of $P_{\beta_i}$ for some $\beta_i\in \{1,\dots,k\}$. 
For $x\in V(J)$, in bag $B_x$, replace each instance of the vertex of $H_i$ corresponding to $R_{i+1}^-$  by the vertex of $H$ corresponding to $P_{\beta_i}$.

@@

\begin{lem}
	An embedded multigraph $G$ is edge-maximal 1-plane with no faces of length 2 if and only if there is a 2-connected plane graph $G_0$ such that: 
	\begin{compactitem}
		\item each face of $G_0$ is bounded by a 3-cycle or a 4-cycle,
		\item no edge of $G_0$ is in the boundary of two facial 3-cycles,
		\item $G=\widehat{G_0}$.
	\end{compactitem}
\end{lem}

>>>>

\begin{lem}
Say $G_0$ is a 2-connected plane multigraph. Let $T$ be a spanning tree of $G_0$ rooted at some vertex $r$ on the boundary of the outerface of $G_0$. Let $k\in\{1,2,3\}$ and let $C = [A_1,\dots,A_k]$ be a cycle in $G_0$ with $r$ in the exterior of $C$, where $A_1, \dots, A_k$ are bipods in $G_0$. Let $G$ be the plane graph consisting of all the edges and vertices of $G_0$ in $C$ and the interior of $C$. Then $G$ has a partition $\PP$ into tripods with respect to $T$ such that:
\begin{compactitem}
\item $P_1,\dots,P_k\in\PP$,
\item $G / \PP$ has a tree-decomposition $\mathcal{T}$ of width at most 3 such that:
\begin{compactitem}
	\item some bag of $\mathcal{T}$ contains $P_1,\dots,P_k$, and 
	\item for every face $F$ of $G$ there is a bag $B$ of $\mathcal{T}$ such that $V(F)\subseteq \bigcup_{X\in B} X$.
\end{compactitem}
\end{compactitem}
\end{lem}

\begin{proof}
	\note{DW}{Can we always assume $k=3$ by allowing $A_i=\emptyset$? That would clean up the proof.}

@@

Consider $j\in\{1,\dots,m\}$. Let $G_j$ be the subgraph of $G$ consisting of all the edges and vertices of $G$ in $Z_j$ and the interior of $Z_j$. The outerface of $G_j$ is $Z_j$. Observe that for some $i\in\{1,\dots,k\}$, we have
$Z_j=[R_{i,j}^+,R_{i+1,j}^-,S_{i,j}]$, where 
$R_{i,j}^+$ is some subpath of $R_i^+$,
$R_{i+1,j}^-$ is some subpath of $R_{i+1}^-$, and
$S_{i,j}$ is some subpath of $S_i$.
Since $R_i^+$, $R_{i+1}^-$ and $S_i$ are bipods, 
$R_{i,j}^+$, $R_{i+1,j}^-$ and $S_{i,j}$ are bipods. 
By construction, $v_{i-1}$ is not in $G_j$. 
Thus $V(G_j)\setminus V(Z_j) \subsetneq V(G)\setminus V(C)$, and we may apply induction to $G_j$.  Thus $G_j$ has a partition $\PP_j$ into tripods with respect to $T$ such that:
\begin{compactitem}
	\item $R_{i,j}^+, R_{i+1,j}^-, S_{i,j}\in\PP_j$,
	\item the quotient $G_j / \PP_j$ has a tree-decomposition $\mathcal{T}_j=\{B_x:x\in V(D_j)\}$ of width at most 3 such that: 
	\begin{compactitem}
		\item some bag of $\mathcal{T}_j$ contains $R_{i,j}^+, R_{i+1,j}^-, S_{i,j}$, and 
		\item for every face $F$ of $G_j$ there is a bag $B$ of $\mathcal{T}_j$ such that $V(F)\subseteq \bigcup_{X\in B} X$.
	\end{compactitem}
\end{compactitem}

@@

We now construct the desired partition $\PP$ of $G$. 
Let $\PP$ be obtained from $\bigcup_{j\in\{1,\dots,m\}} \PP_j$ by replacing 
$R_{i,j}^+\in\PP_j$ by $P_i$, 
$R_{i+1,j}^-\in\PP_j$ by $P_i$, and
$S_{i,j}\in\PP_j$ by $Y$.
By construction, $\PP$ partitions $V(G)$ into tripods with respect to $T$ and $P_1,\ldots,P_k\in \PP$.

We now construct the desired tree-decomposition of $G/\PP$. For each $j\in\{1,\dots,m\}$, apply the above replacements within the bags of $\mathcal{T}_j$, and let $u_j$ be the node of $D_j$ such that $R_{i,j}^+, R_{i+1,j}^-, S_{i,j} \in B_{u_j}$. Let $D$ be the tree obtained from the disjoint union of $D_1,\dots,D_m$ by adding one new node $u$ adjacent to $u_1,\dots,u_m$. Define $B_u:=\{P_1,P_2,P_3,Y\}$. We claim that $\{B_x:x\in V(D)\}$ is the desired tree-decomposition of $G/\PP$. By construction, every bag $B_x$ has size at most 4.

First we show that for each part $X\in\PP$, the set $\{x\in V(D) : X\in B_x\}$ induces a subtree of $D$. If $X\not\in \{P_1,P_2,P_3,Y\}$ then $X$ is only in bags in some $D_j$, implying $\{x\in V(D) : X\in B_x\}$ induces a subtree of $D$. If $X\in \{P_1,P_2,P_3,Y\}$ then $X$ is in $B_u$, and if some part $X'$ of some $\PP_j$ was replaced by $X$ in the above construction, then $X'$ is in $B_{u_j}$ and $\{x\in V(D_j):X\in B_x\}$ induces a subtree of $D_j$, implying that  $\{x\in V(D) : X\in B_x\}$ induces a subtree of $D$ (since $uu_j\in E(D)$).

We now show that, for every edge $XW$ of $G/\PP$, there is a node $x\in V(D)$ such that $X,W\in B_x$. If $X,W\in \{P_1,P_2,P_3,Y\}$ then $X,W\in B_u$ as desired. Otherwise, without loss of generality, $X\in \PP_j\setminus \{P_1,P_2,P_3,Y\}$ for some $j\in\{1,\dots,m\}$. 
Say $X'\in\PP'_j$ replaced by $X\in\PP_j$. Thus $X\subseteq V(G_j)\setminus V(Z_j)$. Since $Z_j$ separates $V(G_j)\setminus V(Z_j)$ and $V(G)\setminus V(G_j)$ and $XW\in E(G/\PP)$, it must be that $W\in \PP_j$ also, implying that $XW\in B_x$ for some node $x\in V(D_j)$. 
This concludes the proof that $(B_x:x\in V(D))$ is the desired tree-decomposition  of $G/\PP$.

Every face $F$ of $G$ is a face of some $G_j$ or is $F$ \note{DW}{Avoid using $F$ for two different things}. In the latter case, if $B$ is the bag at $u$ then $V(F)\subseteq \bigcup_{X\in B} X$ since $V(F) \subseteq Y\in B$. In the first case, there is a bag $B$ of $\mathcal{T}_j$ such that $V(F)\subseteq \bigcup_{X\in B} X$. This property is maintained by the above replacements since 
$R_{i,j}^+ \subseteq P_i$, 
$R_{i+1,j}^- \subseteq P_i$, and
$S_{i,j}\subseteq Y$.

%For each non-degenerate $F_i$, and for each node $x\in V(J_i)$, initialise $B_x:= B^i_x$. 
%Recall that vertices of $H_i$ correspond to contracted paths in  $\PP_i$.
%Each internal path in $\PP_i$ also lies  in $\PP$.
%Each external path $P$ in $\PP_i$ is a subpath of $P_j$ for some $j\in\{1,\dots,k \}$ or is one of the paths among $Q'_1, Q'_2, Q'_3$. 
%For each such path $P$, for every $x\in V(J)$, in bag $B_x$, 
%replace each instance of the vertex of $H_i$ corresponding to $P$ by the vertex of $H$ corresponding to the path among $P_1,\ldots, P_k, Q'_1,\ldots,Q'_3$ 
%that contains $P$.
%This completes the description of $(B_x : x\in V(J))$. 
%By construction, $|B_x|\leq 9$ for every $x\in V(J)$.

%We construct $\PP$ as before. Initialise $\PP:=\{P_1,\ldots,P_k,Y\}$. Then, for $i\in \{1,2,3\}$, each tripod in  $\PP_i$ is either fully contained in $F_i$ or it is \emph{internal} with none of its vertices in $F_i$. Add all these internal tripods in  $\PP_i$ to $\PP$. By construction, $\PP$ partitions $V(G)$ into tripods. The graph $H:= G/ \PP$ is planar since $G$ is planar and each tripod in $\PP$ induces a connected subgraph of $G$.

%Next we produce the tree-decomposition $(B_x:x\in V(J))$ of $H$ that satisfies the requirements of the lemma.  Let $J$ be the tree obtained from the disjoint union of $J_1$, $J_2$ and $J_3$ by adding one new node $u$ adjacent to $u_1$, $u_2$ and $u_3$. Let $B_u$ be the set of at most four vertices of $H$ corresponding to $Y,P_1,\dots,P_k$. For $i\in\{1,2,3\}$ and for each node $x\in V(J_i)$, initialise $B_x:= B^i_x$.

%As in  the proof of \cref{NearTriang},  the resulting structure,  $(B_x:x\in V(J))$, is not yet a tree-decomposition of $H$ since some bags may  contain vertices of $H_i$ that are not necessarily vertices of $H$. Note that unlike in  \cref{NearTriang} this does not only include elements of $\PP_i$  that are contained in $F$. In particular, $I_i$ is also not an element of $\PP$ and thus does not correspond to a vertex of $H$.   We remedy this  as follows. For $x\in V(J)$, in bag $B_x$, replace each instance of the vertex of $H_i$ corresponding to $I_i$ by the vertex of $H$ corresponding to $Y$. Similarly, by construction, $R_i^+$ is a subgraph of $P_{\alpha_i}$ for some $\alpha_i\in \{1,\dots,k\}$.  For $x\in V(J)$, in bag $B_x$, replace each instance of the vertex of $H_i$ corresponding to $R_i^+$  by the vertex of $H$ corresponding to $P_{\alpha_i}$. Finally, $R_{i+1}^-$ is a subgraph of $P_{\beta_i}$ for some $\beta_i\in \{1,\dots,k\}$. For $x\in V(J)$, in bag $B_x$, replace each instance of the vertex of $H_i$ corresponding to $R_{i+1}^-$  by the vertex of $H$ corresponding to $P_{\beta_i}$.

@@

\begin{lem}
	An embedded multigraph $G$ is edge-maximal 1-plane with no faces of length 2 if and only if there is a 2-connected plane multigraph $G_0$ such that: 
	\begin{compactitem}
		\item each face of $G_0$ is bounded by a 3-cycle or a 4-cycle,
		\item no edge of $G_0$ is in the boundary of two facial 3-cycles,
		\item $G=\widehat{G_0}$.
	\end{compactitem}
\end{lem}
